---
title: "A. digitata plastome assembly"
author: "Oketch Fredrick Onyango"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the required packages

```{r echo=FALSE, message=FALSE}
pacman::p_load(
  tidyverse,
  viridis,
  reshape2,
  ggpubr
)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(tidyr)
library(patchwork)
library(ggbreak)
library(ggdendro)
library(factoextra)
library(ggplotify)
library(gridExtra)
```


## ANALYSIS OF ANI and PCA 

```{r, echo=FALSE, message=FALSE}
library(ggplot2)
library(ggdendro)
library(factoextra)
library(patchwork)
library(ggplotify)
library(gridExtra)

# Load ANI data
source("scripts/ani.R")

ani_dist <- 100 - ani

hc <- hclust(
  as.dist(ani_dist),
  method = "average"
)

p98 <- plot(
  hc,
  xlab = "",
  sub = "",
  ylab = "ANI distance (100 − ANI)"
)

p98

# Load PCA script
source("scripts/pca.R")

# Hierarchical clustering
hc <- hclust(
  as.dist(100 - ani),
  method = "average"
)

# Convert dendrogram to ggplot
p_ani <- as.ggplot(~plot(
  hc,
  xlab = "",
  sub = "",
  ylab = "ANI distance (100 − ANI)"
))



# PCA plot
p_pca <- fviz_pca_biplot(
  pca,
  repel = TRUE,
  col.var = "contrib",
  col.ind = "red",
  geom.ind = c("point", "text"),
  pointsize = 3,
  labelsize = 4,
  addEllipses = FALSE
) +
  stat_ellipse(
    aes(x = pca$x[,1], y = pca$x[,2]),
    type = "euclid",
    level = 3.5,
    color = "black",
    linewidth = 0.5
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )


# Combine
combined <- p_ani + p_pca +
  plot_layout(ncol = 2) +
  plot_annotation(tag_levels = "a")

combined

ggsave(
  filename = "figures_r/Figure 2.pdf",
  plot = combined,
  device = cairo_pdf,
  width = 14,
  height = 8,
  units = "in",
  dpi = 300
)
```
## Adansonia digitata synteny and Inverted Repeart boundary analysis


```{r}
library(genoPlotR)
library(ape)

# Load genebank files
gbk_files <- list.files(
  "data_1",
  pattern = "\\.gb$",
  full.names = TRUE
)
dna_segs <- lapply(gbk_files, read_dna_seg_from_genbank)
names(dna_segs) <- gsub(
  "\\.gb$",
  "",
  basename(gbk_files)
)


# Load phylogenetic tree
tree <- read.tree("data_1/adansonia_tree.treefile")
if (!all(tree$tip.label %in% names(dna_segs))) {
  stop("Tree tip labels do not match GenBank file names")
}
dna_segs <- dna_segs[tree$tip.label]


# Convert tree for genoPlotR
phylo_obj <- newick2phylog(write.tree(tree))


# Build pairwise comparison
species <- names(dna_segs)
comparisons <- lapply(1:(length(species) - 1), function(i) {
  sp1 <- species[i]
  sp2 <- species[i + 1]
  f1 <- paste0(
    "data_2/blast_results/",
    sp1,
    "_vs_",
    sp2,
    ".blast"
  )
  
  f2 <- paste0(
    "data_2/blast_results/",
    sp2,
    "_vs_",
    sp1,
    ".blast"
  )
  
  if (file.exists(f1)) {
    
    read_comparison_from_blast(f1)
    
  } else if (file.exists(f2)) {
    
    read_comparison_from_blast(f2)
    
  } else {
    
    stop(
      paste(
        "Missing BLAST comparison file:",
        sp1,
        "vs",
        sp2
      )
    )
  }
})


# Colour gene by function
func_colors <- c(
  "NDH"       = "#E41A1C",
  "PSI"       = "#377EB8",
  "PSII"      = "#4DAF4A",
  "ATP"       = "#FF7F00",
  "RPO"       = "#984EA3",
  "Ribosomal" = "#A65628",
  "PET"       = "#F781BF",
  "RBC"       = "#FFFF33",
  "Other"     = "#999999"
)

assign_colors <- function(seg) {
  n <- tolower(seg$name)
  cols <- ifelse(
    grepl("^ndh", n), func_colors["NDH"],
    ifelse(
      grepl("^psa", n), func_colors["PSI"],
      ifelse(
        grepl("^psb", n), func_colors["PSII"],
        ifelse(
          grepl("^atp", n), func_colors["ATP"],
          ifelse(
            grepl("^rpo", n), func_colors["RPO"],
            ifelse(
              grepl("^rpl|^rps", n), func_colors["Ribosomal"],
              ifelse(
                grepl("^pet", n), func_colors["PET"],
                ifelse(
                  grepl("^rbc", n), func_colors["RBC"],
                  func_colors["Other"]
                )
              )
            )
          )
        )
      )
    )
  )
  
  seg$col  <- cols
  seg$fill <- cols
  return(seg)
}
dna_segs <- lapply(dna_segs, assign_colors)


#Colour synteny links
comparisons <- lapply(comparisons, function(comp) {
  comp$col <- apply_color_scheme(
    comp$per_id,
    "grey90",
    rng = c(10, 100)
  )
  comp$fill <- comp$col
  return(comp)
})


# Annotate the top genome only
gene_lengths <- dna_segs[[1]]$end -
  dna_segs[[1]]$start

# Name only genes longer than 700 bp
selected <- gene_lengths > 700

annots <- annotation(
  x1   = middle(dna_segs[[1]])[selected],
  text = dna_segs[[1]]$name[selected],
  rot  = 90,
  col  = "black"
)

# Create empty annotation for other genomes
annots_list <- c(
  list(annots),
  vector("list", length(dna_segs) - 1)
)

# Export figure as PDF
cairo_pdf(
  "figures_r/Figure 3.pdf",
  width  = 22,
  height = 10
)

# Increase top margin to prevent annotation overlap
par(mar = c(4, 4, 10, 4))

plot_gene_map(
  dna_segs          = dna_segs,
  comparisons       = comparisons,
  annotations       = annots_list,
  tree              = phylo_obj,
  tree_width        = 3,
  annotation_height = 3,
  annotation_cex    = 1,
  dna_seg_labels    = names(dna_segs),
  dna_seg_label_cex = 1.5,
  seg_plot_height   = 1,
  gene_type         = "arrows",
  scale             = TRUE
)

dev.off()
```

# Relative codon usage analysis in comparison with other adansonai species
```{r}
source("scripts/rscu_script.R")

# Save high quality PDF
ggsave(
  "figures_r/Figure 4.pdf",
  p,
  width = 13,
  height = 8,
  units = "in",
  dpi = 300
)

```


## Gene locus identification and characterization-

```{r pressure, echo=FALSE}
# Load and transform the required data
gene <- read_csv("data/gene_coordinates_full.csv") |>
  janitor::clean_names()

sgenes <- gene |>
  filter(copy_status == "Single")

dgenes <- gene |>
  filter(copy_status == "Duplicate") |>
  select(gene) |>
  unique()

lsc <- gene |>
  filter(region == "LSC") |>
  select(gene) |> 
  unique()

irb <- gene |>
  filter(region == "IRb") |>
  select(gene) |>
  unique()

ira <- gene |>
  filter(region == "IRa") |>
  select(gene) |>
  unique()

ssc <- gene |>
  filter(region == "SSC") |>
  select(gene) |>
  unique()


uirs <- bind_rows(ira, irb) |>  
  distinct(gene, .keep_all = TRUE)
```
```{r, echo=FALSE, message=FALSE}
# Load data---------------------------------------------------------------------
codon <- read_csv("data/codon_df.csv")

codon$GC12 <- (3 * codon$GC - codon$GC3s) / 2
codon$AT_bias <- codon$A3s / (codon$A3s + codon$T3s)
codon$GC_bias <- codon$G3s / (codon$G3s + codon$C3s)
codon$eNC <- 2 + codon$GC3s + 29 /
  ((codon$GC3s^2) + ((1 - codon$GC3s)^2))
codon$ENC_diff_ratio <- (codon$eNC - codon$Nc)/
  codon$eNC

```


# PLOTS

## Nc–GC3 plot 
```{r, echo=FALSE, message=FALSE}
library(dplyr)
library(ggrepel)
library(ggplot2)

# Select 10 genes with strongest positive codon bias and 5 with weakest
label_genes <- codon |> 
  arrange(desc((eNC-Nc)/eNC)) |> 
  slice_head(n = 10)  |> 
  bind_rows(
    codon  |> 
      arrange((eNC-Nc)/eNC)  |> 
      slice_head(n = 10)
  )

# Create theoretical curve
expected_curve <- data.frame(
  GC3s = seq(0, 1, length.out = 200)
)

expected_curve$Nc <- 2 + expected_curve$GC3s +
  29 / ((expected_curve$GC3s^2) +
          ((1 - expected_curve$GC3s)^2))

# Plot
p3 <- ggplot() +
  geom_line(data = expected_curve,
            aes(x = GC3s, y = Nc),
            color = "red", size = 1,
            linetype = "dashed") +
  
  geom_point(data = codon,
             aes(x = GC3s, y = Nc),
             color = "black",
             size = 1.5, alpha = 0.7) +
  
  geom_text_repel(data = label_genes,
                  aes(x = GC3s, y = Nc,
                      label = Genes),
                  size = 3) +
  
  theme_classic(base_size = 14) +
  labs(x = expression(GC[3]),
       y = "NC")
p3
```


## eNc plot 
```{r, echo=FALSE, message=FALSE}
# Compute density-based smooth CDFs --------------------------------------------
dens_obs <- density(codon$Nc, na.rm = TRUE)
dens_exp <- density(codon$eNC, na.rm = TRUE)

cdf_obs <- cbind(dens_obs$x, cumsum(dens_obs$y) / sum(dens_obs$y))
cdf_exp <- cbind(dens_exp$x, cumsum(dens_exp$y) / sum(dens_exp$y))

# Get starting and ending points
x_min <- min(c(codon$Nc, codon$eNC), na.rm = TRUE)
x_max <- max(c(codon$Nc, codon$eNC), na.rm = TRUE)

# Extend observed curve to start 5 units before lowest ENC
x_start_obs <- x_min - 0
cdf_obs <- rbind(c(x_start_obs, 0), cdf_obs)

# Extend expected curve to start from 0
cdf_exp <- rbind(c(0, 0), cdf_exp)

# Combine into one data frame for ggplot
df_cdf <- data.frame(
  x = c(cdf_obs[,1], cdf_exp[,1]),
  y = c(cdf_obs[,2], cdf_exp[,2]),
  Type = rep(c("Observed ENC", "Expected eNC"),
             each = nrow(cdf_obs))
)

# Perform Kolmogorov–Smirnov test ---
ks_result <- ks.test(codon$Nc, codon$eNC)
p_value <- formatC(ks_result$p.value, format = "e", digits = 2)
D_value <- round(ks_result$statistic, 3)

# Plot with extended lines and classic theme ---
p1 <- ggplot(df_cdf, aes(x = x, y = y, color = Type)) +
  geom_line(size = 0.8) +
  scale_color_manual(values = c("blue", "red")) +
  theme_classic(base_size = 14) +
  labs(
    x = "Value",
    y = "Cumulative Distribution Function",
    subtitle = paste0("K–S Test: D = ", D_value, ",  P = ", p_value)
  ) +
  coord_cartesian(xlim = c(x_start_obs, x_max), ylim = c(0, 1)) +
  theme(
    legend.position = "right",
    legend.title = element_blank(),
    plot.subtitle = element_text(size = 12, face = "italic")
  )

p1
```
## Frequency distribution of eNc values

```{r, echo=FALSE, message=FALSE}
# Define bin width
bin_width <- 0.08

# Create histogram 
hist_data <- hist(
  codon$ENC_diff_ratio,
  breaks = seq(
    floor(min(codon$ENC_diff_ratio,
              na.rm = TRUE)),
    ceiling(max(codon$ENC_diff_ratio,
                na.rm = TRUE)),
    by = bin_width
  ),
  plot = FALSE
)

# Convert to data frame
bin_df <- data.frame(
  Bin_Start = head(hist_data$breaks, -1),
  Bin_End = tail(hist_data$breaks, -1),
  Frequency = hist_data$counts
)

# Calculate percentage
bin_df$Percentage <- (bin_df$Frequency /
                        sum(bin_df$Frequency)) * 100

# Remove bins with zero frequency
bin_df <- subset(bin_df,
                 Frequency > 0)

# Add midpoint for plotting
bin_df$Midpoint <- (bin_df$Bin_Start +
                      bin_df$Bin_End) / 2

# Show cleaned table
bin_df

# Plot the frequency distribution histogram
p2 <- ggplot(bin_df, aes(x = Midpoint,
                         y = Percentage)) +
  geom_bar(stat = "identity",
           fill = "grey50", color = "black",
           width = bin_width * 0.95) +
  #geom_text(aes(label = sprintf("%.1f", Percentage)),
           # vjust = -0.3, size = 3.5) +
  theme_classic(base_size = 14) +
  labs(
    x = expression((eNC - NC)/eNC),
    y = "Frequency"
  ) +
  geom_vline(xintercept = 0,
             linetype = "dashed",
             color = "black")

p2
```

## Neutrality plot with regression equation
### Relationship between GC content at the third codon position

```{r, echo=FALSE, message=FALSE}

p4 <- ggplot(codon, aes(x = GC3s,
                        y = GC12)) +
  geom_point(size = 1.5, color = "black",
             alpha = 0.4) +
  geom_smooth(method = "lm", color = "black",
              se = TRUE) +
  
# Regression line equation
  stat_regline_equation(
    aes(label = paste(..eq.label..,
                      ..rr.label.., sep = "~~~")),
    label.x = 0.2, label.y = 0.6,
    size = 4, parse = TRUE
  ) +
  
# Correlation coefficient and p-value
  stat_cor(
    method = "pearson",
    label.x = 0.2,
    label.y = 0.55,
    size = 4
  )  +
theme(
  axis.text.x = element_text(size = 12, angle = 0, hjust = 1),
  axis.text.y = element_text(size = 12),
  strip.text.y = element_text(size = 12, face = "bold"),
  axis.title  = element_text(size = 12, face = "bold")
) +
  labs(
    x = "GC3s",
    y = "GC12s"
  ) +
  
  theme_classic()
p4
```
```{r}
library(patchwork)
library(ggbreak)
# Cumulative and GC12 graph
graph2 <- ((p3 | p1) /
           (p2 | p4)) +
  plot_annotation(tag_levels = "a")

# Save high quality PDF
ggsave(
  "figures_r/Figure 5.pdf",
  graph2,
  width = 13,
  height = 8,
  units = "in",
  dpi = 300
)

graph2
```

## PR2 Plot

```{r, echo=FALSE, message=FALSE}
p5 <- ggplot(codon, aes(x = GC_bias, y = AT_bias)) +

  # reference lines
  geom_vline(xintercept = 0.5,
             linetype = "dashed",
             linewidth = 0.6) +
  geom_hline(yintercept = 0.5,
             linetype = "dashed",
             linewidth = 0.6) +

  # points
  geom_point(size = 2.2, alpha = 0.7, color = "darkblue") +

  # labels
  labs(
    x = "G3 / (G3 + C3)",
    y = "A3 / (A3 + T3)"
  ) +

  # base theme
  theme_bw() +

  # theme refinements (clean, manuscript-ready)
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(linewidth = 0.4,
                                colour = "grey40"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13, face = "bold")
  ) +

  guides(fill = "none")

p5
```
#COA_ANALYSIS
```{r, echo=FALSE, message=FALSE}
library(ade4)
library(ggplot2)
library(ggrepel)
library(ggbreak)

coa1 <- read.table("data/coa.txt",
                   sep = "\t",          
                   stringsAsFactors = FALSE,
                   header = TRUE)

codons <- read.table("data/pca.txt",
                     header = TRUE,
                     sep = "",  # auto-detect whitespace
                     stringsAsFactors = FALSE,
                     fill = TRUE) |>
  mutate(cond_end = str_extract(Codon, "[ACUG]$"))


scale_factor <- 8   # adjust visually if needed
codons$Axis1 <- codons$Axis1 * scale_factor
codons$Axis2 <- codons$Axis2 * scale_factor


p7 <- ggplot(codons, aes(x = Axis1,
                         y = Axis2,
                         color = cond_end)) +

  # reference axes
  geom_hline(yintercept = 0,
             linetype = "dashed",
             linewidth = 0.6) +
  geom_vline(xintercept = 0,
             linetype = "dashed",
             linewidth = 0.6) +

  # points
  geom_point(size = 2.2, alpha = 0.8) +

  # gene labels
  #geom_text_repel(
   ##  size = 3,
    #### ) +

  # labels
  labs(
    x = "Axis 1 (17.48%)",
    y = "Axis 2 (10.51%)"
  ) +

  # base theme
  theme_bw() +

  # theme refinements
 theme(
  panel.grid = element_blank(),
  panel.border = element_rect(linewidth = 0.5, colour = "grey40"),
  axis.text = element_text(size = 12),
  axis.title = element_text(size = 13, face = "bold"),
  plot.title = element_text(face = "bold", hjust = 0.5),

  legend.position = c(0.98, 0.02),
  legend.justification = c(1, 0),
  legend.direction = "vertical",
  legend.background = element_rect(fill = "white", colour = "white"),

  legend.title = element_blank(),             
  legend.text = element_text(face = "bold")    
)

p7

```

```{r, echo=FALSE, message=FALSE}
# Calculate A3(A3+T3) and G3(G3+C3) ratio---------------------------------------
aa_ratio <- codon
aa_ratio$A3s <- aa_ratio$A3s/(aa_ratio$A3s+aa_ratio$T3s)
mean_A3_ratio <- mean(aa_ratio$A3s, na.rm = TRUE)
sd_A3_ratio   <- sd(aa_ratio$A3s, na.rm = TRUE)

cat("Mean A3/(A3+T3):", round(mean_A3_ratio, 3), "\n")
cat("Standard Deviation:", round(sd_A3_ratio, 3), "\n")


gg_ratio <- codon
gg_ratio$G3s <- gg_ratio$G3s/(gg_ratio$G3s + gg_ratio$C3s)
mean_G3_ratio <- mean(gg_ratio$G3s, na.rm = TRUE)
sd_G3_ratio   <- sd(gg_ratio$G3s, na.rm = TRUE)

cat("Mean G3/(G3+C3):", round(mean_G3_ratio, 3), "\n")
cat("Standard Deviation:", round(sd_G3_ratio, 3), "\n")
```

```{r, echo=FALSE, message=FALSE}
library(patchwork)
library(ggbreak)

graph3 <- (p5 | p7)+
  plot_annotation(tag_levels = 'a')

# Save high quality PDF
ggsave(
  "figures_r/Figure 6.pdf",
  graph3,
  width = 13,
  height = 8,
  units = "in",
  dpi = 300
)

graph3
```

# Nucleotide diversity
```{r, echo=FALSE, message=FALSE}

# Genic regions
nu_gene <- read.table("data/gene_sort_as_cp_order.txt",
                     header = TRUE,
                     sep = "",            
                     stringsAsFactors = FALSE,
                     fill = TRUE) |>
  filter(Pi != 0)

nu_gene$pi <- as.numeric(nu_gene$Pi)

loc <- read.table("data/gene_location.txt",
                      header = TRUE,
                      sep = "",            
                      stringsAsFactors = FALSE,
                      fill = TRUE)


nu_gene <- merge(nu_gene,
                 loc[, c("Gene", "Location")],
                 by = "Gene", all.x = TRUE) |>
  unique() 


# intergenic region

nu_igs <- read.table("data/IGS_sort_as_cp_order.txt",
                      header = TRUE,
                      sep = "",            
                      stringsAsFactors = FALSE,
                      fill = TRUE) |>
  filter(Pi > 0.000890) |>
  unique()

nu_igs$pi <- as.numeric(nu_igs$Pi)

# Ensure IGS is treated as a factor in the original order
nu_igs <- nu_igs |> 
  mutate(IGS = factor(IGS, levels = unique(IGS)))

data1 <- nu_gene |>
  rename("Locus" = "Location") |> 
  mutate(role = "Protein-coding genes") |>
  unique()

data2 <- nu_igs |>
  rename("Gene" = "IGS") |> 
  mutate(role = "Intron and Intergenic region") |>
  unique()

box2 <- bind_rows(data1, data2) |>
  unique() |>
  dplyr::filter(pi <= 0.15)
```

```{r}
# Nucleotide diversity
mean_pi <- mean(nu_gene$pi, na.rm = TRUE)
sd_pi   <- sd(nu_gene$pi, na.rm = TRUE)


p19 <- ggplot(nu_gene, aes(x = Gene, y = pi, color = Location)) +
  geom_point(size = 2) +
  geom_hline(yintercept = mean_pi, linetype = "dashed") +
  geom_text_repel(
    data = subset(nu_gene, pi > mean_pi),
    aes(label = Gene),
    size = 4,
    nudge_y = 0.001,      # slightly above points
    angle = 90,
    max.overlaps = Inf
  ) +
  scale_color_manual(values = c("LSC" = "black", "SSC" = "red", "IR" = "blue")) +
  labs(x = "Gene", y = "Nucleotide Diversity (π)", color = "Genome Region") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

p19
```

```{r}
# Nucleotide diversity
mean_pie <- mean(nu_igs$pi, na.rm = TRUE)
sd_pie   <- sd(nu_igs$pi, na.rm = TRUE)



p21 <- ggplot(nu_igs, aes(x = IGS, y = pi, color = Locus)) +
  geom_point(size = 2) +
  geom_hline(yintercept = mean_pie, linetype = "dashed") +
  geom_text_repel(
    data = subset(nu_igs, pi > mean_pie),
    aes(label = IGS),
    size = 4,
    nudge_y = 0.001,      # slightly above points
    angle = 90,
    max.overlaps = Inf
  ) +
  scale_color_manual(values = c("LSC" = "black", "SSC" = "red", "IR" = "blue")) +
  labs(x = "Gene", y = "Nucleotide Diversity (π)", color = "Genome Region") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
p21
```

```{r}
# make data
df1 <- nu_gene |>
  rename("Locus" = "Location")

df3 <- nu_igs |>
  rename("Gene" = "IGS")

box1 <- bind_rows(df1, df3) |>
  unique() |>
  dplyr::filter(pi <= 0.15)
  

# Add a 'Type' column to distinguish the two datasets
box1_mod <- box1 |>  mutate(Type = "Locus")
box2_mod <- box2 |>  mutate(Type = "Role")

# Combine the datasets
combined_box <- bind_rows(
  box1_mod |>  rename(Label = Locus),
  box2_mod |>  rename(Label = role)
)


combined_box <- combined_box %>%
  mutate(Label = recode(Label,
                        "Protein-coding genes" = "CDS",
                        "Intron and Intergenic region" = "IIR"))


# Plot
p22 <- ggplot(combined_box, aes(x = Label, y = pi, fill = Type)) +
  geom_boxplot(color = "black") +
  theme_classic() +
  labs(x = "", y = "Nucleotide diversity") +
  scale_fill_manual(values = c("Locus" = "grey70", "Role" = "#3F4403"), guide = "none") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  ) +
  ylim(0, max(combined_box$pi) * 1.05)

p22

```

## Merging the nucleotide diversity graphs
```{r, echo=FALSE, message=FALSE}
library(patchwork)

clean_theme <- theme(
  plot.margin = margin(0, 0, 0, 0),
  axis.title = element_text(size = 13, color = "black"),
  axis.text  = element_text(size = 13, color = "black")
)

# Apply clean theme and remove legend for p19
p19 <- p19 + clean_theme + theme(legend.position = "none")
p21 <- p21 + clean_theme
p22 <- p22 + clean_theme
# p20 removed

# Layout
top_row    <- p19 + p22 + plot_layout(widths = c(7, 3))
bottom_row <- p21  # p20 removed

panel <- top_row / bottom_row +
  plot_layout(heights = c(1, 1)) +
  plot_annotation(tag_levels = "a")

panel

# Save high quality PDF
ggsave(
  "figures_r/Figure 7.pdf",
  panel,
  width = 16,
  height = 10,
  units = "in",
  dpi = 300
)
```



# RNA editing sites

```{r}
library(tidyverse)

# Load data
edit <- read_tsv(
  "data/editing_sites.txt",
  na = c("", "NA"),
  show_col_types = FALSE,
  trim_ws = TRUE
)

# Standardize column names
colnames(edit) <- c(
  "gene",
  "no",
  "step",
  "base_pos",
  "aa_pos",
  "codon_pos",
  "base_change",
  "codon_change",
  "aa_change",
  "label",
  "silent"
)

# Fill missing gene names downward
edit <- edit |>
  fill(gene)

# Remove duplicated rows
edit <- edit |>
  distinct()

# Number of editing sites
n_total <- nrow(edit)

n_total


# Number of genes and editing sites
gene_counts <- edit |>
  count(gene, sort = TRUE) |>
  mutate(percent = n / sum(n) * 100)

gene_counts


conv_counts <- edit |>
  count(base_change) |>
  mutate(percent = n / sum(n) * 100)

codon_counts <- edit |>
  count(codon_pos) |> 
  mutate(percent = n / sum(n) * 100)

aa_counts <- edit |>
  count(aa_change, sort = TRUE)
```

```{r}
library(tidyverse)
library(patchwork)
library(viridis)

# ---------------------------------------------------------
# 1. Editing sites per gene
# ---------------------------------------------------------

p_gene <- gene_counts |>
  filter(n > 6) |>
  ggplot(
    aes(
      x = reorder(gene, n),
      y = n,
      fill = n
    )
  ) +
  geom_col() +
  scale_fill_viridis_c(
    option = "turbo",
    direction = 1
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip() +
  theme_classic() +
  labs(
    x = "Gene",
    y = "Number of editing sites",
    fill = "Editing sites"
  )


# ---------------------------------------------------------
# 2. Base conversion types
# ---------------------------------------------------------

p_conversion <- ggplot(
  conv_counts,
  aes(
    x = base_change,
    y = n,
    fill = n
  )
) +
  geom_col(width = 0.6) +
  scale_fill_viridis_c(
    option = "turbo",
    begin = 0,
  end = 1
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_classic() +
  labs(
    x = "Conversion type",
    y = "Count",
    fill = "Count"
  )


# ---------------------------------------------------------
# 3. Codon position
# ---------------------------------------------------------

p_codon <- ggplot(
  codon_counts,
  aes(
    x = factor(codon_pos),
    y = n,
    fill = n
  )
) +
  geom_col(width = 0.6) +
  scale_fill_viridis_c(
    option = "turbo",
    begin = 0,
  end = 1
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_classic() +
  labs(
    x = "Codon position",
    y = "Editing sites",
    fill = "Editing sites"
  )


# ---------------------------------------------------------
# 4. Amino-acid changes
# ---------------------------------------------------------

p_aa <- aa_counts |>
  slice_max(n, n = 26) |>
  ggplot(
    aes(
      x = reorder(aa_change, n),
      y = n,
      fill = n
    )
  ) +
  geom_col(width = 0.85) +
  scale_fill_viridis_c(
    option = "turbo",
    begin = 0.4,
  end = 1
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip() +
  theme_classic() +
  labs(
    x = "Amino-acid change",
    y = "Count",
    fill = "Count"
  )


# ---------------------------------------------------------
# 5. Silent vs non-silent
# ---------------------------------------------------------

p_silent <- edit |>
  mutate(
    type = ifelse(
      is.na(silent) | silent == "",
      "Non-silent",
      "Silent"
    )
  ) |>
  count(type) |>
  ggplot(
    aes(
      x = type,
      y = n,
      fill = n
    )
  ) +
  geom_col(width = 0.6) +
  scale_fill_viridis_c(
    option = "D",
    direction = 1
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_classic() +
  labs(
    x = "Editing type",
    y = "Count",
    fill = "Count"
  )


# ---------------------------------------------------------
# 6. Distribution along plastome
# ---------------------------------------------------------

p_position <- ggplot(
  edit,
  aes(
    x = base_pos
  )
) +
  geom_histogram(
    binwidth = 200,
    aes(fill = after_stat(count))
  ) +
  scale_fill_viridis_c(
    option = "D",
    direction = 1
  ) +
  theme_classic() +
  labs(
    x = "Nucleotide position",
    y = "Editing count",
    fill = "Editing count"
  )


# ---------------------------------------------------------
# 7. Gene functional groups
# ---------------------------------------------------------

edit <- edit |>
  mutate(
    group = case_when(
      gene %in% c(
        "psaA", "psaB", "psaI", "psaJ",
        "psbB", "psbF", "psbH", "psbJ",
        "psbK", "psbN", "psbT", "psbZ",
        "ycf4"
      ) ~ "Photosystem",
      str_detect(gene, "^ndh") ~ "NDH",
      str_detect(gene, "^atp") ~ "ATP synthase",
      gene %in% c(
        "petA", "petD", "petG", "petL", "ccsA"
      ) ~ "Cytochrome",
      str_detect(gene, "^rps|^rpl") ~ "Ribosomal",
      str_detect(gene, "^rpo") ~ "RNA polymerase",
      gene == "rbcL" ~ "Carbon fixation",
      gene == "accD" ~ "Fatty acid metabolism",
      gene == "matK" ~ "Intron splicing",
      gene == "clpP" ~ "Protein quality control",
      gene == "cemA" ~ "Membrane/transport",
      gene %in% c("ycf1", "ycf2") ~ "Conserved plastid proteins",
      TRUE ~ "Other"
    )
  )


p_group <- edit |>
  count(group) |>
  ggplot(
    aes(
      x = reorder(group, n),
      y = n,
      fill = n
    )
  ) +
  geom_col(width = 0.9) +
  scale_fill_viridis_c(
    option = "D",
    direction = 1
  ) +
  coord_flip() +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_classic() +
  labs(
    x = "Functional group",
    y = "Number of editing sites",
    fill = "Editing sites"
  )

```

# Combine the plots
```{r}
library(patchwork)

design <- "
ABC
DE#
"

combined_editing <-
  p_conversion +
  p_gene +
  p_aa +
  p_codon +
  p_group +
  plot_layout(
    design = design
  ) +
  plot_annotation(
    tag_levels = "A"
  ) &
  theme(
    legend.position = "none"
  )

combined_editing

# Save as high-resolution PDF
ggsave(
  filename = "figures_r/Figure 8.pdf",
  plot = combined_editing,
  width = 12,
  height = 9,
  units = "in",
  device = cairo_pdf,
  dpi = 300
)
```


```{r}
# Selection Analysis - Visualization & Metrics
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(pheatmap)
library(RColorBrewer)


# Load data
df <- readr::read_csv("data/all_models_merged.csv")


# Prepare M0 data
m0 <- df %>%
  filter(model == "M0") %>%
  mutate(
    omega_clean = ifelse(omega >= 900, NA, omega),   
    omega_flag  = ifelse(omega >= 900, "boundary (dS≈0)", 
                         ifelse(omega > 1,
                                ">1 (pos.sel.)",
                                "≤1 (purifying)")),
    mean_dN = as.numeric(mean_dN),
    mean_dS = as.numeric(mean_dS),
    func_group = case_when(
      grepl("^atp", gene) ~ "ATP synthase",
      grepl("^ndh", gene) ~ "NDH complex",
      grepl("^pet", gene) ~ "Electron transport",
      grepl("^psa", gene) ~ "Photosystem I",
      grepl("^psb", gene) ~ "Photosystem II",
      grepl("^rpl", gene) ~ "Ribosomal",
      grepl("^rps", gene) ~ "Ribosomal",
      grepl("^rpo", gene) ~ "RNA polymerase",
      grepl("^rbc", gene) ~ "Carbon fixation",
      grepl("^ycf", gene) ~ "ycf",
      grepl("^mat", gene) ~ "Maturase",
      grepl("^clp", gene) ~ "Protease",
      TRUE                ~ "Other protein-coding genes"
    )
  )


# LRT calculations (M1a vs M2a, M7 vs M8 
lnL_wide <- df %>%
  select(gene, model, lnL) %>%
  filter(model %in% c("M0","M1a","M2a","M7","M8")) %>%
  pivot_wider(names_from = model, values_from = lnL, names_prefix = "lnL_")

lrt <- lnL_wide %>%
  mutate(
    LRT_M1a_M2a = 2 * (lnL_M2a - lnL_M1a),
    p_M1a_M2a   = pchisq(LRT_M1a_M2a, df = 2, lower.tail = FALSE),
    sig_M1a_M2a = p_M1a_M2a < 0.05,
    LRT_M7_M8   = 2 * (lnL_M8 - lnL_M7),
    p_M7_M8     = pchisq(LRT_M7_M8, df = 2, lower.tail = FALSE),
    sig_M7_M8   = p_M7_M8 < 0.05,
    pos_sel_both = sig_M1a_M2a & sig_M7_M8
  ) |>
  pivot_longer(
    cols = c(lnL_M0, lnL_M1a, lnL_M2a, lnL_M7, lnL_M8),
    names_to = "model",
    values_to = "lnL"
  ) 

cat("=== LRT Results: M1a vs M2a ===\n")
lrt %>% filter(sig_M1a_M2a) %>%
  select(gene, LRT_M1a_M2a, p_M1a_M2a) %>%
  arrange(p_M1a_M2a) %>% print()

cat("\n=== LRT Results: M7 vs M8 ===\n")
lrt %>% filter(sig_M7_M8) %>%
  select(gene, LRT_M7_M8, p_M7_M8) %>%
  arrange(p_M7_M8) %>% print()

cat("\n=== Genes significant in BOTH tests ===\n")
lrt %>% filter(pos_sel_both) %>%
  select(gene, LRT_M1a_M2a, p_M1a_M2a, LRT_M7_M8, p_M7_M8) %>%
  print()
```

```{r}
library(ggbeeswarm)
library(ggplot2)
library(dplyr)
library(ggplot2)
library(dplyr)

p_omega <- m0 |> 
  filter(!is.na(omega_clean)) |>
  mutate(func_group = ifelse(func_group %in% c("Protease",
                                               "Maturase",
                                               "ycf",
                                               "Carbon fixation",
                                               "Other protein-coding genes"),
                             "Other protein CDS",
                             func_group)) |> 
  
  ggplot(aes(x = func_group, y = omega_clean, fill = func_group)) +
  geom_violin(trim = FALSE, alpha = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red", linewidth = 0.8) +
  coord_flip() +
  scale_fill_viridis_d(option = "turbo") +
  labs(x = "Functional group", y = "Ka/Ks Ratio", fill = "Functional Group") +
  theme_classic(base_size = 11) +
  theme(legend.position = "none")+
  theme(
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold")
  )

print(p_omega)
```

```{r}
library(tidyverse)
library(grid) 

# -----------------------------
# 1. Input data
# -----------------------------
df <- readr::read_csv("data/all_models_merged.csv")

# Extract lnL by model
lnL_wide <- df %>%
  select(gene, model, lnL) %>%
  filter(model %in% c("M0","M1a","M2a","M7","M8")) %>%
  pivot_wider(names_from = model, values_from = lnL, names_prefix = "lnL_")

lrt <- lnL_wide %>%
  mutate(
    # M1a vs M2a: df = 2
    LRT_M1a_M2a = 2 * (lnL_M2a - lnL_M1a),
    p_M1a_M2a   = pchisq(LRT_M1a_M2a, df = 2, lower.tail = FALSE),
    sig_M1a_M2a = p_M1a_M2a < 0.05,
    # M7 vs M8: df = 2
    LRT_M7_M8   = 2 * (lnL_M8 - lnL_M7),
    p_M7_M8     = pchisq(LRT_M7_M8, df = 2, lower.tail = FALSE),
    sig_M7_M8   = p_M7_M8 < 0.05,
    # Both significant = strong evidence
    pos_sel_both = sig_M1a_M2a & sig_M7_M8
  ) |>
  pivot_longer(
    cols = c(lnL_M0, lnL_M1a, lnL_M2a, lnL_M7, lnL_M8),
    names_to = "model",
    values_to = "lnL"
  ) 


# -----------------------------
# 2. Reshape to long format
# -----------------------------
lrt_data <- lrt 

lrt_long <- data.frame(
  gene = rep(lrt_data$gene, 2),
  
  test = rep(c("M1a_vs_M2a", "M7_vs_M8"),
             each = nrow(lrt_data)),
  
  p = c(lrt_data$p_M1a_M2a,
        lrt_data$p_M7_M8)
)

# -----------------------------
# 3. Compute -log10(p)
# -----------------------------
lrt_long <- lrt_long %>%
  mutate(
    logP = -log10(p)
  )

# -----------------------------
# 4. Rank genes by strongest signal
# -----------------------------
gene_rank <- lrt_long %>%
  group_by(gene) %>%
  summarise(max_logP = max(logP, na.rm = TRUE)) %>%
  arrange(desc(max_logP)) %>%
  mutate(rank = row_number())

# -----------------------------
# 4b. KEEP ONLY TOP 30 GENES
# -----------------------------
top_genes <- gene_rank %>%
  slice_head(n = 30)

# Filter dataset to top 30 genes only
lrt_long <- lrt_long %>%
  filter(gene %in% top_genes$gene)

# Reapply correct ordering
lrt_long$gene <- factor(lrt_long$gene,
                        levels = top_genes$gene)


# Highlight genes under selection
highlight_genes <- c("petA", "ycf1", "rbcL", "rpoC1")

lrt_long <- lrt_long %>%
  mutate(
    group = ifelse(gene %in% highlight_genes,
                   "Genes under selection",
                   "Other genes")
  )

# Plot graph

P_hierachy <- ggplot(lrt_long,
       aes(x = gene, y = logP)) +
  
  # Points
  geom_point(aes(color = test, shape = group),
             size = 2,
             position = position_dodge(width = 0.5)) +
  
  # Significance threshold
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed",
             color = "grey30",
             linewidth = 0.2) +
  
  # Custom colors for model comparison
  scale_color_manual(
    values = c(
      "M1a_vs_M2a" = "blue",
      "M7_vs_M8"   = "red"
    ),
    labels = c(
      "M1a_vs_M2a" = "M1a vs M2a",
      "M7_vs_M8"   = "M7 vs M8"
    )
  ) +
  
  # Labels
  labs(
    x = "Protein Coding Genes",
    y = expression(-log[10](p-value)),
    color = "Model comparison",
    shape = "Gene category"
  ) +
  
  # Theme
  theme_classic(base_size = 14) +
  
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid.minor = element_blank(),
    
    # Bold legend titles
    legend.title = element_text(face = "bold")
  ) +
  theme(
  legend.position = c(0.97, 0.97),
  legend.justification = c(1, 1),

  legend.title = element_text(size = 10, face = "bold"),
  legend.text  = element_text(size = 10),

  legend.key.size = unit(0.35, "cm")
)

print(P_hierachy)
```

```{r}
library(patchwork)

combined_plot <- (p_omega | P_hierachy) +
  
  plot_layout(widths = c(0.5, 0.5)) 

# Save high quality PDF
ggsave(
  "figures_r/Figure 9.pdf",
  combined_plot,
  width = 16,
  height = 8,
  units = "in",
  dpi = 300
)

print(combined_plot)
```

```{r, echo=TRUE, warning=TRUE}

# Phylogenetic tree
library(ape)
library(phytools)
library(stringr)


ml_46 <- read.tree("data/ml_46.txt")
bi_46 <- read.tree("data/bi_46.txt")


# Plot side by side
par(mfrow = c(1, 2),
    mar = c(2, 2, 4, 2))

plotTree(ml_46,
         direction = "rightwards",
         ftype = "i",
         lwd = 1,
         fsize = 0.7)
title("a")

plotTree(bi_46,
         direction = "leftwards",
         ftype = "i",
         lwd = 1,
         fsize = 0.7)

title("b")


pdf("figures_r/ML_BI_tree.pdf",
    width = 14,
    height = 12)

dev.off()
```


